# What is data visualization?

```{r, include = FALSE}
ottrpal::set_knitr_image_path()
```

## Learning Objectives

```{r fig.align = 'center', out.width = "100%", echo = FALSE, fig.alt = "Chapter specific LOs go here"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3015fd53a67_4_176#slide=id.g3015fd53a67_4_176")
```

## What is Data Visualization?

Definition box: Data visualization is the process of making visual representations such as plots or other figures to to show different representations of data, to both explore the data or communicate aspects about the data including trends, patterns, relationship or steps in a process. 

There are two main types of data visualizations: Exploratory and Expository visualizations.

```{r fig.align = 'center', out.width = "100%", echo = FALSE, fig.alt = "Exploratory versus Expository visualizations."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3a4b3643292_0_46#slide=id.g3a4b3643292_0_46")
```

### Exploratory Visualizations

Exploratory Data Analysis (EDA)

### Expository Visualizations

Expository visualizations are polished visualizations prepared for the purpose of sharing. You may share an expository visualization within a preprint, a peer-reviewed publication, on a poster, or as part of a presentation.

Because expository visualizations will be shared widely, you will not always be there to present or explain the visualization to others. Therefore, you want to design your data visualization to **stand alone**. 

With this goal in mind, an important part of creating an expository visualization also includes writing the narrative to go along with it. This narrative may involve an informative title, well-labeled axes, a thorough caption, and alternative text. Note that not all formats require each of these elements to appear alongside the visualization. You won't put a caption on your presentation slide, however you may want to include the caption within the speaker notes.

<details><summary>Specific tips for a high quality narrative</summary>

* Ensure the title states the plot's main conclusion
* Avoid using field specific jargon or obscure acronyms within a title or an axis label unless defined (perhaps within the caption).

</details>  

Additionally, because an expository visualization will be shared widely, the audience will be broad and have varied interests and expertise. If you are giving a semi-frequent presentation within your department or lab group or presenting within a session at a specialized conference, you can expect the audience to be somewhat familiar with the topic. However, no one will know your research like you do. Many others may see the visualization outside of your department or institution -- perhaps even patients or their family members! Therefore, you want to prepare your expository visualizations in a way that they will be accessible to a larger audience.

Creating a good expository visualization will take time. Data visualization is often an iterative process both as you prepare the plot and as you seek and incorporate feedback from your team. Be sure to carefully inspect and validate your visualization before sharing it widely. A common phrase within home maintenance and improvement communities is "Measure twice, cut once". As you prepare your expository visualization, you want to check it twice before you share it the first time.  

## Why is data visualization so important?
