---
title: "Wrapping up"
format:
  html:
    toc: true
    toc-depth: 4
    code-annotations: hover
---

## Learning Objectives

```{r los, fig.align = "center", echo = FALSE, out.width="100%", fig.alt = "This chapter will demonstrate how to iteratively construct a data visualization, evaluate whether a data visualization can be improved to enhance its clarity, accessibility, or accuracy, and to discuss ways to hone data visualization skills"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3bc553a32e9_0_0#slide=id.g3bc553a32e9_0_0")
```

## Iterative Example

To summarize all of the considerations and best practices discussed in the chapters before this, the first section of this final chapter will walk through an example, step-by-step, of iteratively building up an expository data visualization. Each step or iteration showcases a different view of the same data as we refine how to effectively communicate the message we want to convey for the audience.

Throughout this example exercise, you will see several types of specially-colored boxes. In particular, there are four to which you should pay special attention.

One box highlights potential ethical issues and best practices to avoid those issues.

<div class = "ethics">

In this box, you'll find reminders of the ethical considerations you should be particularly aware of while building data visualizations as well as steps you can take to ensure best practices.

</div>

Another box highlights potential accessibility issues and best practices to avoid those issues.

<div class = "accessibility">

In this box, you'll find reminders of the accessibility considerations you should be particularly aware of while building data visualizations. These include considerations for specific audience groups as well as considerations for increasing clarity for the audience in general. These boxes will include steps you can take to ensure best practices.

</div>

One box highlights specific thought questions or prompts for you to consider as you work through this example.

::: {.emphasis_block}

<div class = "reflection">

In this box, you'll find a specific prompt which poses a thought question about the material for you to consider.

</div>

<details><summary>Example answer in click to expand section</summary>

In this click to expand section, you'll find a possible answer to the thought question.

</details>

:::

The final box often follows the thought question box, but not always. It highlights where within the course various topics were covered.

<div class = "notice">

In this box, you'll find information connecting topics to where they were covered in the course in case you want to review those topics.

</div>


{{< include _example_data_description.qmd >}}

:::{.emphasis_block}

<div class = "reflection">
What variables (specifically what types of variables -- categorical or numerical) are part of this dataset as described?
</div>

<details open><summary>Variables in the candy dataset</summary>

This means that the candy dataset (at least given the way we have worked with or "wrangled" it) has a few main variables of interest:

1. Names of the candy
    * A **categorical** variable
    * 76 different values
2. Proportion of respondents who reported joy
    * A **numerical** variable
    * Specifically a *continuous* numerical variable
    * Bounded between 0 and 1.0
    * 76 different values - one for each candy
3. Proportion of respondents who reported despair
    * A **numerical** variable
    * Specifically a *continuous* numerical variable
    * Bounded between 0 and 1.0
    * 76 different values - one for each candy

There are additional variables such as responses to demographic questions as well as the proportion of respondents who reported indifference or didn't respond at all. However, to simplify the message we wish to communicate with our data visualization, we won't directly use these variables for this example.

</details>

:::

<div class="notice">
To review variables and different types of data, visit [Chapter 3](https://hutchdatascience.org/DataViz_Considerations/03_data.html#data-types)
</div>

### Goal

Our goal is to visualize a ranking for the most liked candies in 2017 according to this dataset.

```{r goalImage, fig.align = "center", echo = FALSE, out.width="100%", fig.alt = "This is an iterative example putting it all together. Given a 2017 candy dataset that asks respondents to report their feelings of joy despair or indifference, our goal is to visualize a ranking of the most liked candies"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3bb275f27df_0_6#slide=id.g3bb275f27df_0_6")
```

### Choosing a plot type

::: {.emphasis_block}

<div class="reflection">
Given the data that we have to work with and the goal of the visualization, what type of plot would you use?
</div>

<details><summary>Comparing and contrasting plot types for this visualization</summary>

Note that respondents did not simply rank the candy based on preference and we instead have two numerical values for each candy. Therefore we should probably consider plots types beyond the conventional ranking plots described in Chapter 4.

**A more conventional approach**:

A bar plot or even a stacked bar plot could work if we computed a single score (proportion of Joy - proportion of Despair perhaps), but with 76 candies, we would need to arrange the bars based on the magnitude or size and just focus on the bars on the extremes. The visualization would likely appear pretty cramped overall if we tried to show every candy.

  * Here's an example of an analysis that utilized bar plots: [https://github.com/phoebewong/candy-hierarchy-2017](https://github.com/phoebewong/candy-hierarchy-2017). Notice, for this example they focused on a couple specific candies which gets around the "cramped" visualization concern.

  * Similar to a bar plot is a lollipop chart (seems fitting for a candy dataset): [https://x.com/ttrodrigz/status/923582440937021440](https://x.com/ttrodrigz/status/923582440937021440). This example uses a log ratio rather than a difference for each candy, sorts based on magnitude, and then focuses on the extremes as expected.

**A more unconventional approach**:

A scatter plot is constructed with two numerical variables and is often used to show a correlation. We might expect our data to have some sort of a correlation because likely candy that is greatly enjoyed by many won't be despised by many. And conversely, candy that is despised by many probably won't be greatly enjoyed. The only way to see if this expectation is true is to explore the data and just plot it. So that's what we'll do for the first iteration.

</details>

:::

<div class="notice">
To review different conventional plot types, visit [Chapter 4](https://hutchdatascience.org/DataViz_Considerations/04_graph_info.html#standard-graph-types).
</div>

### Building the plot

The first few iterations will be an exploratory analysis of the data where we don't focus on aesthetics, but rather:

* what the data looks like
* if patterns match our expectations
* if our chosen plot style will work for us
* perhaps even hypothesis or idea generation

Then the remaining iterations will transition into polishing the visualization in order to prepare an expository data visualization.

<div class="notice">
To review the differences between exploratory and expository data visualizations, visit [Chapter 2](https://hutchdatascience.org/DataViz_Considerations/02_what_why_intro.html#what-is-data-visualization).
</div>


#### Plot the data

{{< include _example_iter1_plot_data.qmd >}}

::: {.emphasis_block}

<div class="reflection">
So far, this is just using 2 of the 3 main variables we identified. How can we use the third variable?
</div>

<details><summary>Plotting the candy names</summary>
We can use a labeling function to add the candy names to the plot, labeling specific points.
</details>
:::

#### Labeling all points

{{< include _example_iter2_label_all_points.qmd >}}

::: {.emphasis_block}

<div class="reflection">
What do you notice after observing the labels? Do you have any ideas on new variables or a takeaway message?
</div>

<details><summary>Takeaways from the candy name labeling</summary>
It looks like chocolate candy (Twix, Kit Kat, 100 Grand Bar...) is among the most liked candy. It could be beneficial to create a new categorical variable that represents if each candy is chocolate or not.
</details>
:::

#### Distinguishing groups

{{< include _example_iter3_use_color.qmd >}}

::: {.emphasis_block}

<div class="reflection">
What do you observe about the group coloring?
</div>

<details><summary>Takeaways from the group coloring</summary>
It looks even more convincing that chocolate candy is among the most liked candy -- a strong pattern that could be our takeaway message.
</details>
:::

#### Using shape with color

{{< include _example_iter4_use_shape.qmd >}}

<div class = "accessibility">

When using color to distinguish groups, an important accessibility step is to use shape as a redundant way to distinguish groups, especially if the color palette being used isn't necessarily color vision deficiency friendly. This redundancy increases accessibility for those with color vision deficiency.

</div>


#### Adjusting point size, transparency, and color

{{< include _example_iter5_adjust_points.qmd >}}

<div class = "accessibility">

When using color to distinguish groups, to increase accessibility, use tools to check your color palette to see if the colors can be distinguished by those with color vision deficiency.

</div>

#### Ordering of axes to promote readability

{{< include _example_iter6_order_axes.qmd >}}

#### Adding quadrant delineations

{{< include _example_iter7_quadrants.qmd >}}


#### Labeling quadrants

{{< include _example_iter8_quadrant_labels.qmd >}}

{{< include _example_iter8_quadrant_emojis.qmd >}}

#### Distinguishing quadrant labels

{{< include _example_iter9_color_labels.qmd >}}

{{< include _example_iter9_simplify.qmd >}}


#### Informative title

{{< include _example_iter10_title.qmd >}}

#### Labeling extreme and border points

{{< include _example_iter11_label_extreme_points.qmd >}}

#### Fixing the coordinate system aspect ratio

{{< include _example_iter12_aspect_ratio.qmd >}}

<div class = "ethics">

When axes have the same limits, fixing the aspect ratio to 1:1 is the most ethical approach to avoid distorting data.

</div>

#### Simplifying labels and increasing text size

{{< include _example_iter13_axis_legend_labels.qmd >}}

#### The plot!

```{r finalPlot, fig.align = "center", echo = FALSE, out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3be338b6b90_0_146#slide=id.g3be338b6b90_0_146")
```

But we're not done yet....

#### The caption!

```{r plotCaption, fig.align = "center", echo = FALSE, out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3bb275f27df_0_20#slide=id.g3bb275f27df_0_20")
```

:::{.emphasis_block}

<div class = "reflection">

What might be a good caption for this expository data visualization? Try writing one! Focus on providing a detailed description of what the plot shows that will help the plot to tell its story and stand alone.

</div>

<details><summary>A possible caption</summary>

**Chocolate is among the most liked candy in 2017.** A survey of potential trick or treaters in 2017 by the University of British Columbia asked how survey respondents felt when they received certain candies. Here each candy is plotted based on the fraction of respondents which viewed that candy with Joy vs the fraction of respondents which viewed that candy with Despair. The upper right corner of the plot represents the most liked candies (lots of joy, little despair), while the bottom left corner represents the least liked candies (more despair, less joy). Candy color and shape is used to separate chocolate candy (brown triangles) from non-chocolate candy (blue circles). Chocolate candy seems to be among the most liked candy with any full-sized candy bar having the most joy and least despair reported.

</details>

:::

#### Going further

The dataset is an open dataset, meaning that it is available for you to explore and visualize on your own!

Perhaps you are interested in whether there's a difference in rankings depending on the sex of respondent or some other demographics. Do males prefer chocolate over non-chocolate candy? Does the age or location of the respondent impact the rankings?

Perhaps you are interested in how the proportion of indifference affects these rankings. Is there a design element that can be added to this plot to visualize that specifically?

Perhaps you'd like to try to plot this with a completely different plot type! Whatever your question -- go for it! Practice is one of the best ways to improve your data visualization skills.

## Keep Practicing!

With your data within your research projects, keep practicing thorough exploration prior to intentional and careful data visualization, making design choices to increase the clarity and accessibility of your visualizations for a wide audience.

In addition, more open ended practice with open source datasets can be found by browsing data visualization communities and challenges such as

1. [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday)

2. [Posit Plotnine challenges](https://github.com/has2k1/plotnine/discussions/categories/2024-plotnine-contest)

3. [bioviz challenges](http://biovis.net/2026/). Note that these bioviz challenges are more advanced, often describing working with complex multi-omic data and sometimes suggesting that participants should sketch possible visualizations rather than perfectly polishing expository visualization.

As mentioned within the Field Specific Visualizations chapter, data visualization is commonly used within bioinformatics research to validate models. Researchers may wonder how well their model performs. To answer that question, they may run either experimental or simulated data with known "truth" through the model and compare the model's predictions with the known "truth". Researchers may then wonder why the model is messing up in spot where the prediction doesn't line up with the observed truth. To answer this, researchers explore specific examples where the known truth and predictions differ and look for patterns.

Two examples of using data visualization to validate models within the bioinformatic literature include:

1. A model (`Xpresso`) which predicts gene expression from gene sequence [@Agarwal_Shendure_2020]. The authors ran the model using sequences of genes with experimentally known expression values and compared the model predictions with the known / observed expression. Figure 3A is an example of this comparison / validation. Other subpanels within the figure explore where the predictions diverge from the observed expression. The authors [provide the data related to this figure](https://www.cell.com/cms/10.1016/j.celrep.2020.107663/attachment/8f2732a1-4b4d-4c39-aae6-5742b2b946af/mmc2.xlsx). Consider recreating Figure 3A or rethink an entirely  different or partially different approach for comparing the predictions and the observed expression values.

2. A model (`rhapsodi`) was designed to impute the missing genotypes within sparse single gamete DNA sequencing data [@Carioscia_etal_2022]. As a result of imputing the missing genotypes, the donor haplotypes are phased and recombination break points are discovered. Benchmarking of the model's performance was done using simulated data. Specifically the authors simulated fully known genotype sequences for many gametes from donor haplotypes (keeping track of recombination locations). Then sparsity was introduced through further simulation into these genotype sequences. The model was provided the simulated sparse inputs and the model outputs could be compared with the fully known simulated data. Figure 2 illustrates much of this benchmarking with Figure 2—figure supplement 5 focusing on looking for patterns where truth and predictions differed, specifically for the recombination break points. The [data used to produce that figure is available on the associated GitHub](https://github.com/mccoy-lab/transmission-distortion/blob/main/plotting/supp_recomb/fn_fp_tall_df.Rdata). Consider exploring the dataset to look for patterns much like the supplemental figure did.

## In Summary

* Aim for understanding and readability for a broad audience over complexity.
* When you are working with your data and selecting your plot, think about what message you want to communicate.
* Iterate and ask for feedback.
* Carefully check visualizations against data and expectations to ensure accuracy.
* Be careful to avoid common data distortions or unintentional takeaways.

## Checklist

This checklist contains reminders of considerations and steps you should take while building your expository data visualization in order to minimize its complexity, enhance its clarity and accessibility, and assess its accuracy.

{{< include _summary_checklist.qmd >}}
