::: {.panel-tabset}

##### Description

```{r iter1Plot, fig.align = "center", echo = FALSE, out.width="100%", fig.alt = "Plotting the feelings dataset in a basic way with fraction of respondents who reported Joy on the x axis and fraction of respondents who reported despair on the y axis. The data appears to be inversely correlated."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3be338b6b90_0_9#slide=id.g3be338b6b90_0_9")
```

This first iteration is an exploratory data analysis step. We're not worried about making the plot look pretty yet. We just want to look at the data. Primarily we want to see whether or not the relationship that we expect between the joy and despair variables holds true.

##### R code

```{r}
#| label: iter1R
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for plotting a scatter plot"

##proportion of Joy on x-axis
##proportion of Despair on the y-axis
##geom_point for scatter plot
##set xlim and ylim to 0,1 since our data is bounded that way
##theme_bw() is a quick way to adjust theme to make it simpler
##using labs to set quick axis labels
## setting overall text size just for example purposes

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR
            )
         ) +
  geom_point() +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_bw() +
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR") +
  theme(text = element_text(size = 14))
```
Note that normally in a first iteration of exploratory data analysis, we might not set the axis labels (they'll default to the column name of whatever is being plotted.), and we might not adjust the text size. Both of these are done here to add clarity to this example.


##### Python code

```{python}
#| label: iter1Python
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Python code for plotting a scatter plot"


mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')                                                #<1>
fig1, ax1 = plt.subplots()
ax1.scatter(just_candy_props['likeness'], just_candy_props['dislikeness'])
ax1.set_xlim(0,1)
ax1.set_ylim(0,1)
ax1.minorticks_on()                                                                #<1>
ax1.grid(which='major',                                                            #<1>
         linestyle='-', linewidth='0.5',                                           #<1>
         color='grey', alpha=0.7)                                                  #<1>
ax1.grid(which='minor',                                                            #<1>
         linestyle=':', linewidth='0.3',                                           #<1>
         color='grey', alpha=0.5)                                                  #<1>
ax1.set_xlabel('Fraction of respondents who reported JOY')
ax1.set_ylabel('Fraction of respondents who reported DESPAIR')
plt.show()
plt.close()
```

1. Replicating the `theme_bw` style of ggplot

Note that normally in a first iteration of exploratory data analysis, we might not set the axis labels (Python won't add any labels) and we might not adjust the text size. Both of these are done here to add clarity to this example.

:::
