::: {.panel-tabset}

##### Description

```{r iter3Plot, fig.align = "center", echo = FALSE, out.width="100%", fig.alt = "This iteration adds color to distinguish groups, specifically chocolate versus not chocolate groups. Light blue represents chocolate, while light pink represents candy that is not chocolate. We see a fairly clear separation with chocolate on one extreme and non-chocolate on the other with mixing of the groups in the middle."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3be338b6b90_0_39#slide=id.g3be338b6b90_0_39")
```

This third iteration, uses color to distinguish the data points according to whether or not they represent a chocolate candy. We've removed the labeling for now (and will add it back later -- labeling fewer or specific candies only). Note that we're assuming that any full sized candy bar is chocolate since the majority of full sized candy bars tend to have chocolate.

##### R code

```{r}
#| label: iter3R
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for adding color to distinguish groups"

#removed code related to the labeling for now

to_plot$Chocolate <- "No"                   #<4>
to_plot$Chocolate[c(12, 21, 25, 27, 33, 43, #<4>
                    45, 46, 48, 49, 50, 51, #<4>
                    53, 54, 55, 56, 57, 58, #<4>
                    59, 60, 61, 62, 63, 64, #<4>
                    65, 66, 67, 68, 69, 70, #<4>
                    71, 72, 73, 74, 75, 76 )] <- "Yes" #<4>

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR,
             color = Chocolate              #<5>
            )
         ) +
  geom_point() +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_bw() +
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR") +
  theme(text = element_text(size = 14))
```

4. Create a new categorical variable
5. Specify that color should use that new categorical variable

##### Python code

```{python}
#| label: iter3Python
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Python code for adding color to distinguish groups"

import numpy as np

labels = np.full(just_candy_props.shape[0], 'No', dtype='U3')
labels[np.array([12, 21, 25, 27, 33, 43,
       45, 46, 48, 49, 50, 51,
       53, 54, 55, 56, 57, 58,
       59, 60, 61, 62, 63, 64,
       65, 66, 67, 68, 69, 70,
       71, 72, 73, 74, 75, 76])-1] = "Yes"
colors = np.where(labels == "Yes", True, False) .astype(int)

just_candy_props = just_candy_props.assign(Chocolate=labels, colors=colors)

mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')
fig, ax = plt.subplots()
scatter = ax.scatter(just_candy_props['likeness'], just_candy_props['dislikeness'], c = just_candy_props['colors'])
ax.set_xlim(0,1)
ax.set_ylim(0,1)
ax.minorticks_on()
ax.grid(which='major', linestyle='-', linewidth='0.5', color='grey', alpha=0.7)
ax.grid(which='minor', linestyle=':', linewidth='0.3', color='grey', alpha=0.5)
ax.set_xlabel('Fraction of respondents who reported JOY')
ax.set_ylabel('Fraction of respondents who reported DESPAIR')

ax.legend(handles=list(scatter.legend_elements()[0]), labels=list(just_candy_props['Chocolate'].unique()), title = "Chocolate")
plt.show()
plt.close()
```

:::
