::: {.panel-tabset}

##### Description

```{r iter11Plot, fig.align = "center", echo = FALSE, out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3be338b6b90_0_92#slide=id.g3be338b6b90_0_92")
```
For the eleventh iteration of the plot, we want to add back some candy labels. We'll focus on labeling

  * the extremes
    * most liked (quadrant I -- upper right)
    * least liked (quadrant III -- lower left)
  * some of the transition candies where chocolate and non-chocolate candies overlap a bit

##### R code

```{r}
#| label: iter11R
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for labeling the extreme points"

to_plot <- to_plot %>%
  mutate(label = "")

to_plot[nrow(to_plot), "label"] <- "Any full-sized bar"                                         #<34>
to_plot[which(str_detect(to_plot$column_name, "Peanut Butter")), "label"] <- "Reese's PB Cups"  #<34>
to_plot[nrow(to_plot)-2, "label"] <- to_plot[nrow(to_plot)-2, "column_name"]                    #<34>
to_plot[nrow(to_plot)-3, "label"] <- to_plot[nrow(to_plot)-3, "column_name"]                    #<34>
to_plot[nrow(to_plot)-4, "label"] <- to_plot[nrow(to_plot)-4, "column_name"]                    #<34>
to_plot[which(to_plot$column_name == "Nerds"), "label"] <- "Nerds"                    #<35>
to_plot[which(to_plot$column_name == "Skittles"), "label"] <- "Skittles"              #<35>
to_plot[which(to_plot$column_name == "Starburst"), "label"] <- "Starburst"            #<35>
to_plot[which(to_plot$column_name == "Sweet Tarts"), "label"] <- "Sweet Tarts"        #<35>
to_plot[1, "label"] <- "Candy that is the stuff given out for free at restaurants"          #<36>
to_plot[2, "label"] <- "Gum from baseball cards"                                            #<36>
to_plot[which(str_detect(to_plot$column_name, "marshmallow")), "label"] <- "Circus peanuts" #<36>

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR,
             label = label,                                                           #<37>
             color = Chocolate,
             shape = Chocolate
            )
         ) +
  geom_point(size = 3,
             alpha = 0.6
            ) +
  xlim(0, 1) +
  ylim(1, 0) +
  theme_bw() +
  geom_hline(aes(yintercept=0.5), linetype = 'dashed') +
  geom_vline(aes(xintercept=0.5), linetype = 'dashed') +
  annotate("richtext",
           label = "<span style='color: #CB2C2C;'>High</span> Joy<br>
                    <span style='color: #409AA6;'>Low</span> Despair",
           x = 0.955,
           y = 0.025,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  annotate("richtext",
           label = "<span style='color: #CB2C2C;'>High</span> Joy<br>
                    <span style='color: #CB2C2C;'>High</span> Despair",
           x = 0.955,
           y = 0.975,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  annotate("richtext",
           label = "<span style='color: #409AA6;'>Low</span> Joy<br>
                    <span style='color: #CB2C2C;'>High</span> Despair",
           x = 0.045,
           y = 0.975,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  annotate("richtext",
           label = "<span style='color: #409AA6;'>Low</span> Joy<br>
                    <span style='color: #409AA6;'>Low</span> Despair",
           x = 0.045,
           y = 0.025,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  scale_color_manual(values = c("Yes" = "#543210",
                                "No" = "#00008B")) +
  geom_text_repel(show.legend = FALSE, max.overlaps = 16) +                           #<38>
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR",
       title = "<b><span style='color: #543210;'>Chocolate</span> is among the most liked candies in 2017</b>") +
  theme(text = element_text(size = 14),
        plot.title = element_markdown(hjust = 0.5))

```

34. set the display label in a variable for the candies liked the most
35. set the display label for the candies on the border between the two groups of candies
36. set the display label for the candies that are least liked
37. set the aesthetic for labels to point to the new variable for display labels
38. use `geom_text_repel` to display those labels and handle any overlaps

Note that the code is labeling points after visual inspection to manually decide what should be labeled. A more robust approach could use the quadrant cutoff values to select which points should be labeled.

##### Python code

```{python}
#| label: iter11Python
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Python code for labeling the extreme points"

from adjustText import adjust_text

just_candy_props["label"] = ""

just_candy_props.loc[just_candy_props.index[-1], "label"] = "Any full-sized bar"
just_candy_props.loc[just_candy_props.index.str.contains("Peanut Butter", regex=False), "label"] = "Reese's PB Cups"
just_candy_props.loc[just_candy_props.index[-3], "label"] = just_candy_props.index[-3]
just_candy_props.loc[just_candy_props.index[-4], "label"] = just_candy_props.index[-4]
just_candy_props.loc[just_candy_props.index[-5], "label"] = just_candy_props.index[-5]

for candy in ["Nerds", "Skittles", "Starburst", "Sweet Tarts"]:
    just_candy_props.loc[just_candy_props.index == candy, "label"] = candy

just_candy_props.loc[just_candy_props.index[0], "label"] = "Candy that is the stuff given out for free at restaurants"
just_candy_props.loc[just_candy_props.index[1], "label"] = "Gum from baseball cards"

just_candy_props.loc[just_candy_props.index.str.contains("marshmallow", regex=False), "label"] = "Circus peanuts"

mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')
fig, ax = plt.subplots()
ax.axhline(y=0.5, linestyle = 'dashed', color = 'black')
ax.axvline(x=0.5, linestyle = 'dashed', color = 'black')
scatter_choc = ax.scatter(just_candy_props['likeness'][mask_choc],
                          just_candy_props['dislikeness'][mask_choc],
                          c = just_candy_props['colors'][mask_choc],
                          marker = np.unique(just_candy_props['shapes'][mask_choc]).item(),
                          label = "Yes",
                          alpha = 0.6)
scatter_nc = ax.scatter(just_candy_props['likeness'][mask_nc],
                        just_candy_props['dislikeness'][mask_nc],
                        c = just_candy_props['colors'][mask_nc],
                        marker = np.unique(just_candy_props['shapes'][mask_nc]).item(),
                        label = "No",
                        alpha = 0.6)
ax.set_xlim(0,1)
ax.set_ylim(1,0)
ax_text(x = 0.88, y = 0.1,
        s='<High> Joy\n<Low> Despair',
        highlight_textprops=[{"color": '#CB2C2C'},
                             {"color": '#409AA6'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)


ax_text(x = 0.88, y = 0.875,
        s = '<High> Joy\n<High> Despair',
        highlight_textprops=[{"color": '#CB2C2C'},
                             {"color": '#CB2C2C'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)

ax_text(x = 0.125,  y = 0.875,
        s = '<Low> Joy\n<High> Despair',
        highlight_textprops=[{"color": '#409AA6'},
                             {"color": '#CB2C2C'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)

ax_text(x = 0.125, y = 0.1,
        s = '<Low> Joy\n<Low> Despair',
        highlight_textprops=[{"color": '#409AA6'},
                             {"color": '#409AA6'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)
ax.minorticks_on()
ax.grid(which='major',
        linestyle='-', linewidth='0.5',
        color='grey', alpha=0.7)
ax.grid(which='minor',
        linestyle=':', linewidth='0.3',
        color='grey', alpha=0.5)
ax.set_xlabel('Fraction of respondents who reported JOY')
ax.set_ylabel('Fraction of respondents who reported DESPAIR')
fig_text(x = 0.5, y = 0.9,
         s = '<Chocolate> is among the most liked candies in 2017',
         highlight_textprops = [{"color": "#543210"}],
         ha = 'center',
         va = 'bottom',
         fontweight = 'bold')
texts = [plt.text(just_candy_props['likeness'].iloc[i],
                  just_candy_props['dislikeness'].iloc[i],
                  just_candy_props['label'].iloc[i],                             #<56>
                  fontsize = 6,
                  color = just_candy_props['colors'].iloc[i])                    #<56>
          for i in range(len(just_candy_props))]
adjust_text(texts,                                                               #<56>
  avoid_points = True,
  arrowprops=dict(arrowstyle='-', shrinkA = 5.0))
ax.legend(title="Chocolate", loc = "center right")
plt.show()
plt.close()

```

56. Using adjust_text like in iteration 2 just with a different column of data for the display labels and adding a color for the labels

Note that Python uses 0-based indexing, so the wrangling steps to set display labels use indices offset by one compared to the R code wrangling steps.

:::
