::: {.panel-tabset}

##### Description

```{r iter5Plot, fig.align = "center", echo = FALSE, out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3be338b6b90_0_45#slide=id.g3be338b6b90_0_45")
```

This fifth iteration is another polishing step. It includes changes controlling how the points appear:

  * size
  * transparency
  * color.

For the plot produced from R, we want to increase the point size. For the plot produced from Python, the point size is good already.

For the plots from both R and Python, we want to increase the point transparency to better separate or show data points that are plotted in similar locations.

We also want to set a specific color scheme rather than relying on default colors. Changing the color palette can be both an accessibility consideration and an aesthetic choice. Aesthetically, we're choosing a brown color to represent the chocolate group. The blue for non-chocolate candy appears to be distinguishable from the brown for those with color vision deficiency according to the palette checker, but it is still a good practice to keep both color and shape as a means to distinguish the two groups.


##### R code

```{r}
#| label: iter5R
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for adjusting the point size, transparency, and color"

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR,
             color = Chocolate,
             shape = Chocolate
            )
         ) +
  geom_point(size = 3,                                                           #<7>
             alpha = 0.6                                                         #<8>
            ) +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_bw() +
  scale_color_manual(values = c("Yes" = "#543210",                               #<9>
                                "No" = "#00008B")) +                             #<9>
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR") +
  theme(text = element_text(size = 14))

```

7. This addition sets the size of the points
8. This addition sets the transparency or opacity of the points so that overlapping points are semi-transparent.
9. This addition manually sets a color scheme using hex codes - brown for chocolate group and dark blue for non-chocolate group

##### Python code

```{python}
#| label: iter5Python
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Python code for adjusting the point transparency and color"

just_candy_props['colors'] = np.where(just_candy_props['Chocolate'] == "Yes", "#543210", "#00008B") #<20>

mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')
fig, ax = plt.subplots()
scatter_choc = ax.scatter(just_candy_props['likeness'][mask_choc],                                   #<21>
                          just_candy_props['dislikeness'][mask_choc],                                #<21>
                          c = just_candy_props['colors'][mask_choc],                                 #<21>
                          marker = np.unique(just_candy_props['shapes'][mask_choc]).item(),          #<21>
                          label = "Yes",                                                             #<21>
                          alpha = 0.6)                                                               #<21>
scatter_nc = ax.scatter(just_candy_props['likeness'][mask_nc],                                       #<21>
                        just_candy_props['dislikeness'][mask_nc],                                    #<21>
                        c = just_candy_props['colors'][mask_nc],                                     #<21>
                        marker = np.unique(just_candy_props['shapes'][mask_nc]).item(),              #<21>
                        label = "No",                                                                #<21>
                        alpha = 0.6)                                                                 #<21>
ax.set_xlim(0,1)
ax.set_ylim(0,1)
ax.minorticks_on()
ax.grid(which='major',
        linestyle='-', linewidth='0.5',
        color='grey', alpha=0.7)
ax.grid(which='minor',
        linestyle=':', linewidth='0.3',
        color='grey', alpha=0.5)
ax.set_xlabel('Fraction of respondents who reported JOY')
ax.set_ylabel('Fraction of respondents who reported DESPAIR')


ax.legend(title="Chocolate")
plt.show()
plt.close()
```

20. Changing the color column in the pandas dataframe to no longer be 1/0 integers but now be hex codes for specific colors. Chocolate rows will be #543210
21. We can remove the vmin and vmax since we are setting a specific color for each scatter call. We also add an alpha to increase the transparency of the points


:::
