::: {.panel-tabset}

##### Description

```{r iter9bPlot, fig.align = "center", echo = FALSE, out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3be338b6b90_0_120#slide=id.g3be338b6b90_0_120")
```

For the plot produced with R, we can simplify the labels for the quadrants by removing the border. The outlines are unnecessary/extraneous marks and could draw the attention of viewers before other elements. We want to be judicious with the geometric marks we use and simplify where we can.

For the plot produced with Python, we can use a pre-existing Python library to simplify the code needed to make the plot.

##### R code

```{r}
#| label: iter9bR
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for simplifying quadrant labels"

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR,
             color = Chocolate,
             shape = Chocolate
            )
         ) +
  geom_point(size = 3,
             alpha = 0.6
            ) +
  xlim(0, 1) +
  ylim(1, 0) +
  theme_bw() +
  geom_hline(aes(yintercept=0.5), linetype = 'dashed') +
  geom_vline(aes(xintercept=0.5), linetype = 'dashed') +
  annotate("richtext",
           label = "<span style='color: #CB2C2C;'>High</span> Joy<br>
                    <span style='color: #409AA6;'>Low</span> Despair",
           x = 0.955,
           y = 0.025,
           fill = NA,                                                   #<29>
           label.color = NA,                                            #<30>
           label.padding = grid::unit(rep(0, 4), "pt")                  #<31>
          ) +
  annotate("richtext",
           label = "<span style='color: #CB2C2C;'>High</span> Joy<br>
                    <span style='color: #CB2C2C;'>High</span> Despair",
           x = 0.955,
           y = 0.975,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  annotate("richtext",
           label = "<span style='color: #409AA6;'>Low</span> Joy<br>
                    <span style='color: #CB2C2C;'>High</span> Despair",
           x = 0.045,
           y = 0.975,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  annotate("richtext",
           label = "<span style='color: #409AA6;'>Low</span> Joy<br>
                    <span style='color: #409AA6;'>Low</span> Despair",
           x = 0.045,
           y = 0.025,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  scale_color_manual(values = c("Yes" = "#543210",
                                "No" = "#00008B")) +
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR") +
  theme(text = element_text(size = 14))

```

29. Makes the label background transparent
30. Removes the box outline from around the label
31. Removes padding

##### Python code

```{python}
#| label: iter9bPython
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Simplified Python code for distinguishing quadrant labels"

from highlight_text import HighlightText, ax_text, fig_text                      #<50>

mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')
fig, ax = plt.subplots()
ax.axhline(y=0.5, linestyle = 'dashed', color = 'black')
ax.axvline(x=0.5, linestyle = 'dashed', color = 'black')
scatter_choc = ax.scatter(just_candy_props['likeness'][mask_choc],
                          just_candy_props['dislikeness'][mask_choc],
                          c = just_candy_props['colors'][mask_choc],
                          marker = np.unique(just_candy_props['shapes'][mask_choc]).item(),
                          label = "Yes",
                          alpha = 0.6)
scatter_nc = ax.scatter(just_candy_props['likeness'][mask_nc],
                        just_candy_props['dislikeness'][mask_nc],
                        c = just_candy_props['colors'][mask_nc],
                        marker = np.unique(just_candy_props['shapes'][mask_nc]).item(),
                        label = "No",
                        alpha = 0.6)
ax.set_xlim(0,1)
ax.set_ylim(1,0)
ax_text(x = 0.88, y = 0.1,                                                       #<51>
        s='<High> Joy\n<Low> Despair',                                           #<51>
        highlight_textprops=[{"color": '#CB2C2C'},                               #<51>
                             {"color": '#409AA6'}],                              #<51>
        fontsize = 10,                                                           #<51>
        ha = 'center',                                                           #<51>
        va = 'center',                                                           #<51>
        textalign = 'center',                                                    #<51>
        ax = ax)                                                                 #<51>


ax_text(x = 0.88, y = 0.875,
        s = '<High> Joy\n<High> Despair',
        highlight_textprops=[{"color": '#CB2C2C'},
                             {"color": '#CB2C2C'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)

ax_text(x = 0.125,  y = 0.875,
        s = '<Low> Joy\n<High> Despair',
        highlight_textprops=[{"color": '#409AA6'},
                             {"color": '#CB2C2C'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)

ax_text(x = 0.125, y = 0.1,
        s = '<Low> Joy\n<Low> Despair',
        highlight_textprops=[{"color": '#409AA6'},
                             {"color": '#409AA6'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)
ax.minorticks_on()
ax.grid(which='major',
        linestyle='-', linewidth='0.5',
        color='grey', alpha=0.7)
ax.grid(which='minor',
        linestyle=':', linewidth='0.3',
        color='grey', alpha=0.5)
ax.set_xlabel('Fraction of respondents who reported JOY')
ax.set_ylabel('Fraction of respondents who reported DESPAIR')


ax.legend(title="Chocolate", loc = "center right")
plt.show()
plt.close()

```

50. Using the highlight_text library rather than our complex function
51. Using the `ax_text` function from the highlight_text library for the upper right label

:::
