::: {.panel-tabset}

##### Description

```{r iter12Plot, fig.align = "center", echo = FALSE, out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3bb275f27df_0_40#slide=id.g3bb275f27df_0_40")
```

For the twelfth iteration, we want to ensure that we are fairly displaying the data rather than distorting it. Our axes are both constrained with data that can only take on values between 0 and 1. Therefore, we want to ensure that our axes are the same length. We can control the aspect ratio to make sure this is the case.

##### R code

```{r}
#| label: iter12R
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for fixing the coordinate ratio"

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR,
             label = label,
             color = Chocolate,
             shape = Chocolate
            )
         ) +
  geom_point(size = 3,
             alpha = 0.6
            ) +
  xlim(0, 1) +
  ylim(1, 0) +
  theme_bw() +
  geom_hline(aes(yintercept=0.5), linetype = 'dashed') +
  geom_vline(aes(xintercept=0.5), linetype = 'dashed') +
  annotate("richtext",
           label = "<span style='color: #CB2C2C;'>High</span> Joy<br>
                    <span style='color: #409AA6;'>Low</span> Despair",
           x = 0.955,
           y = 0.025,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  annotate("richtext",
           label = "<span style='color: #CB2C2C;'>High</span> Joy<br>
                    <span style='color: #CB2C2C;'>High</span> Despair",
           x = 0.955,
           y = 0.975,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  annotate("richtext",
           label = "<span style='color: #409AA6;'>Low</span> Joy<br>
                    <span style='color: #CB2C2C;'>High</span> Despair",
           x = 0.045,
           y = 0.975,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  annotate("richtext",
           label = "<span style='color: #409AA6;'>Low</span> Joy<br>
                    <span style='color: #409AA6;'>Low</span> Despair",
           x = 0.045,
           y = 0.025,
           fill = NA,
           label.color = NA,
           label.padding = grid::unit(rep(0, 4), "pt")
          ) +
  scale_color_manual(values = c("Yes" = "#543210",
                                "No" = "#00008B")) +
  geom_text_repel(show.legend = FALSE, max.overlaps = 16) +
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR",
       title = "<b><span style='color: #543210;'>Chocolate</span> is among the most liked candies in 2017</b>") +
  theme(text = element_text(size = 14),
        plot.title = element_markdown(hjust = 0.5)) +
  coord_fixed(ratio = 1)                                   #<39>

```

39. By using the `coord_fixed` function we can set a fixed aspect ratio, in this case we want a square ratio of 1 because the axes have equal limits

##### Python code

```{python}
#| label: iter12Python
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Python code for fixing the coordinate ratio"

mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')
fig, ax = plt.subplots()
ax.axhline(y=0.5, linestyle = 'dashed', color = 'black')
ax.axvline(x=0.5, linestyle = 'dashed', color = 'black')
scatter_choc = ax.scatter(just_candy_props['likeness'][mask_choc],
                          just_candy_props['dislikeness'][mask_choc],
                          c = just_candy_props['colors'][mask_choc],
                          marker = np.unique(just_candy_props['shapes'][mask_choc]).item(),
                          label = "Yes",
                          alpha = 0.6)
scatter_nc = ax.scatter(just_candy_props['likeness'][mask_nc],
                        just_candy_props['dislikeness'][mask_nc],
                        c = just_candy_props['colors'][mask_nc],
                        marker = np.unique(just_candy_props['shapes'][mask_nc]).item(),
                        label = "No",
                        alpha = 0.6)
ax.set_xlim(0,1)
ax.set_ylim(1,0)
ax_text(x = 0.88, y = 0.1,
        s='<High> Joy\n<Low> Despair',
        highlight_textprops=[{"color": '#CB2C2C'},
                             {"color": '#409AA6'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)


ax_text(x = 0.88, y = 0.875,
        s = '<High> Joy\n<High> Despair',
        highlight_textprops=[{"color": '#CB2C2C'},
                             {"color": '#CB2C2C'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)

ax_text(x = 0.125,  y = 0.875,
        s = '<Low> Joy\n<High> Despair',
        highlight_textprops=[{"color": '#409AA6'},
                             {"color": '#CB2C2C'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)

ax_text(x = 0.125, y = 0.1,
        s = '<Low> Joy\n<Low> Despair',
        highlight_textprops=[{"color": '#409AA6'},
                             {"color": '#409AA6'}],
        fontsize = 10,
        ha = 'center',
        va = 'center',
        textalign = 'center',
        ax = ax)
ax.minorticks_on()
ax.grid(which='major',
        linestyle='-', linewidth='0.5',
        color='grey', alpha=0.7)
ax.grid(which='minor',
        linestyle=':', linewidth='0.3',
        color='grey', alpha=0.5)
ax.set_xlabel('Fraction of respondents who reported JOY')
ax.set_ylabel('Fraction of respondents who reported DESPAIR')
fig_text(x = 0.5, y = 0.9,
        s = '<Chocolate> is among the most liked candies in 2017',
        highlight_textprops = [{"color": "#543210"}],
        ha = 'center',
        va = 'bottom',
        fontweight = 'bold')
texts = [plt.text(just_candy_props['likeness'].iloc[i],
                  just_candy_props['dislikeness'].iloc[i],
                  just_candy_props['label'].iloc[i],
                  fontsize = 6,
                  color = just_candy_props['colors'].iloc[i])
          for i in range(len(just_candy_props))]
adjust_text(texts,
  avoid_points = True,
  arrowprops=dict(arrowstyle='-', shrinkA = 5.0))
ax.legend(title="Chocolate", loc = "center right")
ax.set_aspect('equal')                                                          #<57>
plt.show()
plt.close()

```

57. Setting the aspect of the axes to be equal, the same as setting the aspect ratio to 1

:::
