::: {.panel-tabset}

##### Description

```{r iter8Plot, fig.align = "center", echo = FALSE, out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3be338b6b90_0_71#slide=id.g3be338b6b90_0_71")
```

We want to add some labels to the quadrants in this plot. Viewers need to a quick way to know the significance of where the data points are with respect to the dashed lines.

##### R code

```{r}
#| label: iter8R
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for labeling quadrants"

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR,
             color = Chocolate,
             shape = Chocolate
            )
         ) +
  geom_point(size = 3,
             alpha = 0.6
            ) +
  xlim(0, 1) +
  ylim(1, 0) +
  theme_bw() +
  geom_hline(aes(yintercept=0.5), linetype = 'dashed') +
  geom_vline(aes(xintercept=0.5), linetype = 'dashed') +
  annotate("text",                                        #<17>
           label = "High Joy\nLow Despair",               #<17>
           x = 0.955,                                     #<17>
           y = 0.025                                      #<17>
          ) +                                             #<17>
  annotate("text",                                        #<18>
           label = "High Joy\nHigh Despair",              #<18>
           x = 0.955,                                     #<18>
           y = 0.975                                      #<18>
          ) +                                             #<18>
  annotate("text",                                        #<19>
           label = "Low Joy\nHigh Despair",               #<19>
           x = 0.045,                                     #<19>
           y = 0.975                                      #<19>
          ) +                                             #<19>
  annotate("text",                                        #<20>
           label = "Low Joy\nLow Despair",                #<20>
           x = 0.045,                                     #<20>
           y = 0.025                                      #<20>
          ) +                                             #<20>
  scale_color_manual(values = c("Yes" = "#543210",
                                "No" = "#00008B")) +
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR") +
  theme(text = element_text(size = 14))
```

17. Adding labels to the upper right quadrant (I)
18. Adding labels to the lower right quadrant (IV)
19. Adding labels to the lower left quadrant (III)
20. Adding labels to the upper left quadrant (II)

##### Python code

```{python}
#| label: iter8Python
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Python code for labeling quadrants"

mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')
fig, ax = plt.subplots()
ax.axhline(y=0.5, linestyle = 'dashed', color = 'black')
ax.axvline(x=0.5, linestyle = 'dashed', color = 'black')
scatter_choc = ax.scatter(just_candy_props['likeness'][mask_choc],just_candy_props['dislikeness'][mask_choc], c = just_candy_props['colors'][mask_choc], marker = "^", label = "Yes", alpha = 0.6)
scatter_nc = ax.scatter(just_candy_props['likeness'][mask_nc],just_candy_props['dislikeness'][mask_nc], c = just_candy_props['colors'][mask_nc], marker = "o", label = "No", alpha = 0.6)
ax.set_xlim(0,1)
ax.set_ylim(1,0)
ax.text(0.88, 0.1, "High Joy\nLow Despair", fontsize = 10, ha = 'center', va = 'center')
ax.text(0.88, 0.875, "High Joy\nHigh Despair", fontsize = 10, ha = 'center', va = 'center')
ax.text(0.125, 0.875, "Low Joy\nHigh Despair", fontsize = 10, ha = 'center', va = 'center')
ax.text(0.125, 0.1, "Low Joy\nLow Despair", fontsize = 10, ha = 'center', va = 'center')
ax.minorticks_on()
ax.grid(which='major', linestyle='-', linewidth='0.5', color='grey', alpha=0.7)
ax.grid(which='minor', linestyle=':', linewidth='0.3', color='grey', alpha=0.5)
ax.set_xlabel('Fraction of respondents who reported JOY')
ax.set_ylabel('Fraction of respondents who reported DESPAIR')


ax.legend(title="Chocolate", loc = "center right")
plt.show()
plt.close()
```

:::
