::: {.panel-tabset}

##### Description

```{r iter9aPlot, fig.align = "center", echo = FALSE, out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3be338b6b90_0_125#slide=id.g3be338b6b90_0_125")
```

For the ninth iteration of this plot, we will proceed with the text labels for the quadrants. We want to use color, a pre-attentive attribute, to highlight and synchronize "High" vs "Low".


```{r iter9colorPlot, fig.align = "center", echo = FALSE, out.width="100%"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3bb275f27df_0_0#slide=id.g3bb275f27df_0_0")
```

We'll use red-pink for high and a blue-gray for low. And if we look at all the colors together within the color palette, they still appear to be distinguishable for individuals with various color vision deficiencies.

##### R code

```{r}
#| label: iter9aR
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for distinguishing quadrant labels"

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR,
             color = Chocolate,
             shape = Chocolate
            )
         ) +
  geom_point(size = 3,
             alpha = 0.6
            ) +
  xlim(0, 1) +
  ylim(1, 0) +
  theme_bw() +
  geom_hline(aes(yintercept=0.5), linetype = 'dashed') +
  geom_vline(aes(xintercept=0.5), linetype = 'dashed') +
  annotate("richtext",                                                   #<25>
           label = "<span style='color: #CB2C2C;'>High</span> Joy<br>
                    <span style='color: #409AA6;'>Low</span> Despair",   #<25>
           x = 0.955,
           y = 0.025
          ) +
  annotate("richtext",                                                   #<26>
           label = "<span style='color: #CB2C2C;'>High</span> Joy<br>
                    <span style='color: #CB2C2C;'>High</span> Despair",  #<26>
           x = 0.955,
           y = 0.975
          ) +
  annotate("richtext",                                                   #<27>
           label = "<span style='color: #409AA6;'>Low</span> Joy<br>
                    <span style='color: #CB2C2C;'>High</span> Despair",  #<27>
           x = 0.045,
           y = 0.975
          ) +
  annotate("richtext",                                                   #<28>
           label = "<span style='color: #409AA6;'>Low</span> Joy<br>
                    <span style='color: #409AA6;'>Low</span> Despair",   #<20>
           x = 0.045,
           y = 0.025
          ) +
  scale_color_manual(values = c("Yes" = "#543210",
                                "No" = "#00008B")) +
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR") +
  theme(text = element_text(size = 14))

```

25. Adding color so that "High" in the label is a dark red and "Low" is a light blue in the upper right
26. Adding color so that "High" in the label is a dark red in the lower right
27. Adding color so that "Low" in the label is a light blue and "High" is a dark red in the lower left
28. Adding color so that "Low" in the label is a light blue in the upper left

##### Python code

```{python}
#| label: iter9aPython
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Python code for distinguishing quadrant labels"

from matplotlib import transforms

def multicolor_text(ax, x, y, lines, fontsize=10, hpad=4, vpad=4):               #<44>
    """
    This function was developed collaboratively with Microsoft Copilot

    It draws multi-colored, multi-line text on a Matplotlib axis by rendering each
    word as its own text object. This allows fine-grained control over per-word
    coloring, spacing, and alignment -- similar to richtext annotations in ggplot.

    The function iterates through each line and each word, measuring their widths
    in display coordinates to determine precise horizontal placement. Vertical
    placement is also computed in display coordinates to ensure consistent spacing
    regardless of axis scaling or inversion.

    To mimic ggplot’s default behavior for richtext labels, each line is centered
    relative to a common midpoint (the x data coordinate). This ensures that the
    top line is visually centered over the bottom line, even when the lines have
    different lengths.

    Display coordinates are used internally because Matplotlib’s data coordinates
    cannot guarantee consistent spacing across transforms, DPI settings, or
    inverted axes. Only the final computed positions are converted back into data
    coordinates for drawing.

    Parameters
    ------

    ax: matplotlib.axes.Axes
        The axis on which the annotation will be drawn.

    x, y: float
          Data coordinates where the annotation blocks should be anchored.

    lines: list of lines, where each line is a list of (word, color) tuples.
            Example:
              lines = [
                [("High ", "#CB2C2C"), ("Joy", "black")],
                [("Low ", "#409AA6"), ("Despair", "black")]
              ]

    fontsize: int
              Fontsize for all words

    hpad, vpad: horizontal and vertical padding between words and lines in display points
    """
    fig = ax.figure
    fig.canvas.draw()
    renderer = fig.canvas.get_renderer()

    # Convert the (x, y) data coordinate into display coordinates
    x_disp, y_disp = ax.transData.transform((x, y))

    # First, get a line height from a dummy text
    dummy = ax.text(0, 0, "Ag", fontsize=fontsize, va = "top", ha="left")
    line_height = dummy.get_window_extent(renderer=renderer).height
    dummy.remove()

    # --- Compute the center x for the block ---
    # We use the anchor x as the center
    center_x = x_disp

    # --- Draw each line centered around center_x ---
    for line_idx, line in enumerate(lines):
        # vertical offset DOWN the page in display coords
        y_line = y_disp - (line_height + vpad) * line_idx

        # measure this line's width
        line_width = 0
        for word, color in line:
            t = ax.text(0, 0, word, fontsize=fontsize)
            ex = t.get_window_extent(renderer=renderer)
            t.remove()
            line_width += ex.width + hpad
        line_width -= hpad  # remove trailing padding

        # compute left edge so line is centered
        x_left = center_x - line_width / 2

        # draw the line
        x_offset = 0
        for word, color in line:
            x_word = x_left + x_offset
            x_data, y_data = ax.transData.inverted().transform((x_word, y_line))

            text = ax.text(
                x_data, y_data, word,
                color=color,
                fontsize=fontsize,
                va="top",
                ha="left"
            )

            # measure width for next word
            ex = text.get_window_extent(renderer=renderer)
            x_offset += ex.width + hpad

smile_x = 0.88                                                                  #<45>
smile_y = 0.05                                                                  #<45>
smile_lines = [                                                                 #<45>
    [("High ", "#CB2C2C"), ("Joy", "black")],                                   #<45>
    [("Low ", "#409AA6"), ("Despair", "black")]                                 #<45>
]                                                                               #<45>

smiletear_x = 0.88
smiletear_y = 0.875
smiletear_lines = [
    [("High ", "#CB2C2C"), ("Joy", "black")],
    [("High ", "#CB2C2C"), ("Despair", "black")]
]

sad_x = 0.125
sad_y = 0.875
sad_lines = [
    [("Low ", "#409AA6"), ("Joy", "black")],
    [("High ", "#CB2C2C"), ("Despair", "black")]
]

neutral_x = 0.125
neutral_y = 0.05
neutral_lines = [
    [("Low ", "#409AA6"), ("Joy", "black")],
    [("Low ", "#409AA6"), ("Despair", "black")]
]

mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')
fig, ax = plt.subplots()
ax.axhline(y=0.5, linestyle = 'dashed', color = 'black')
ax.axvline(x=0.5, linestyle = 'dashed', color = 'black')
scatter_choc = ax.scatter(just_candy_props['likeness'][mask_choc],
                          just_candy_props['dislikeness'][mask_choc],
                          c = just_candy_props['colors'][mask_choc],
                          marker = np.unique(just_candy_props['shapes'][mask_choc]).item(),
                          label = "Yes",
                          alpha = 0.6)
scatter_nc = ax.scatter(just_candy_props['likeness'][mask_nc],
                        just_candy_props['dislikeness'][mask_nc],
                        c = just_candy_props['colors'][mask_nc],
                        marker = np.unique(just_candy_props['shapes'][mask_nc]).item(),
                        label = "No",
                        alpha = 0.6)
ax.set_xlim(0,1)
ax.set_ylim(1,0)
fig.canvas.draw()
multicolor_text(ax, smile_x, smile_y, smile_lines, hpad = 0)                    #<46>
multicolor_text(ax, smiletear_x, smiletear_y, smiletear_lines, hpad = 0)        #<47>
multicolor_text(ax, sad_x, sad_y, sad_lines, hpad = 0)                          #<48>
multicolor_text(ax, neutral_x, neutral_y, neutral_lines, hpad = 0)              #<49>
ax.minorticks_on()
ax.grid(which='major',
        linestyle='-', linewidth='0.5',
        color='grey', alpha=0.7)
ax.grid(which='minor',
        linestyle=':', linewidth='0.3',
        color='grey', alpha=0.5)
ax.set_xlabel('Fraction of respondents who reported JOY')
ax.set_ylabel('Fraction of respondents who reported DESPAIR')


ax.legend(title="Chocolate", loc = "center right")
plt.show()
plt.close()
```

44. Defining a function to create multi-color, multi-line annotations for a Matplotlib plot
45. Defining the inputs for the function -- specifically for the upper right quadrant. Repeated for others below.
46. Adding color so that "High" in the label is a dark red and "Low" is a light blue in the upper right
47. Adding color so that "High" in the label is a dark red in the lower right
48. Adding color so that "Low" in the label is a light blue and "High" is a dark red in the lower left
49. Adding color so that "Low" in the label is a light blue in the upper left

:::
