::: {.panel-tabset}

##### Description

```{r iter3Plot, fig.align = "center", echo = FALSE, out.width="100%", fig.alt = "This iteration adds color to distinguish groups, specifically chocolate versus not chocolate groups. Light blue represents chocolate, while light pink represents candy that is not chocolate. We see a fairly clear separation with chocolate on one extreme and non-chocolate on the other with mixing of the groups in the middle."}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3be338b6b90_0_39#slide=id.g3be338b6b90_0_39")
```

This third iteration, uses color to distinguish the data points according to whether or not they represent a chocolate candy. We've removed the labeling for now (and will add it back later -- labeling fewer or specific candies only). Note that we're assuming that any full sized candy bar is chocolate since the majority of full sized candy bars tend to have chocolate.

##### R code

```{r}
#| label: iter3R
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for adding color to distinguish groups"

#removed code related to the labeling for now

to_plot$Chocolate <- "No"                   #<4>
to_plot$Chocolate[c(12, 21, 25, 27, 33, 43, #<4>
                    45, 46, 48, 49, 50, 51, #<4>
                    53, 54, 55, 56, 57, 58, #<4>
                    59, 60, 61, 62, 63, 64, #<4>
                    65, 66, 67, 68, 69, 70, #<4>
                    71, 72, 73, 74, 75, 76 )] <- "Yes" #<4>

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR,
             color = Chocolate              #<5>
            )
         ) +
  geom_point() +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_bw() +
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR") +
  theme(text = element_text(size = 14))
```

4. Create a new categorical variable
5. Specify that color should use that new categorical variable

##### Python code

```{python}
#| label: iter3Python
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Python code for adding color to distinguish groups"

import numpy as np                                                              #<6>

labels = np.full(just_candy_props.shape[0], 'No', dtype='U3')                   #<7>
labels[np.array([12, 21, 25, 27, 33, 43,                                        #<8>
       45, 46, 48, 49, 50, 51,                                                  #<8>
       53, 54, 55, 56, 57, 58,                                                  #<8>
       59, 60, 61, 62, 63, 64,                                                  #<8>
       65, 66, 67, 68, 69, 70,                                                  #<8>
       71, 72, 73, 74, 75, 76])-1] = "Yes"                                      #<8>
colors = np.where(labels == "Yes", True, False).astype(int)                     #<9>

just_candy_props = just_candy_props.assign(Chocolate=labels, colors=colors)     #<10>

mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')
fig, ax = plt.subplots()
scatter = ax.scatter(just_candy_props['likeness'],
                     just_candy_props['dislikeness'],
                     c = just_candy_props['colors'])                            #<11>
ax.set_xlim(0,1)
ax.set_ylim(0,1)
ax.minorticks_on()
ax.grid(which='major',
        linestyle='-', linewidth='0.5',
        color='grey', alpha=0.7)
ax.grid(which='minor',
        linestyle=':', linewidth='0.3',
        color='grey', alpha=0.5)
ax.set_xlabel('Fraction of respondents who reported JOY')
ax.set_ylabel('Fraction of respondents who reported DESPAIR')

ax.legend(handles=list(scatter.legend_elements()[0]),                           #<12>
          labels=list(just_candy_props['Chocolate'].unique()),                  #<12>
          title = "Chocolate")                                                  #<12>
plt.show()
plt.close()
```

6. Using the numpy package for arrays to wrangle the data
7. Creating an array that has the same number of rows as our data where everything is "No" (not chocolate)
8. Changing specific locations in that array to "Yes" (chocolate). Using the same indices as the R code, but subtracting one from all of them since Python uses 0-based indexing or numbering rather than the 1-based R does.
9. Making a new numpy array of Trues and Falses that we convert to an integer (1s and 0s) for the color (which is Matplotlib's preference when it is mapping values to colors)
10. Adding the numpy arrays for color/chocolate grouping as data within our pandas DataFrame
11. Added the color column to the scatter call to specify c or color. Note we are saving the output of this line in a variable for later legend use
12. Creating the legend for color. The Yes/No label from the first numpy array are the labels we display

:::
