::: {.panel-tabset}

##### Description

```{r iter4Plot, fig.align = "center", echo = FALSE, out.width="100%", fig.alt = "This iteration adds shape as a redundant way to distinguish the groups. light blue triangles now represent chocolate candy while light pink circles represent non chocolate"}
ottrpal::include_slide("https://docs.google.com/presentation/d/1fu-KfdN2ldOXB49o9zdpurFbVvl1bOh8cfQ3vu0szk4/edit?slide=id.g3bb275f27df_0_12#slide=id.g3bb275f27df_0_12")
```

This fourth iteration is the first polishing iteration. As discussed in [Chapter 5](https://hutchdatascience.org/DataViz_Considerations/04_how_to_improve_viz.html#accessibility-considerations), to enhance accessibility, we don't want to rely on color alone to distinguish groups. So we'll also use shape to distinguish the groups. This is a bit redundant but can be beneficial to our future audience.

##### R code

```{r}
#| label: iter4R
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "R code for using color and shape together"

to_plot %>%
  ggplot(aes(x = JOY,
             y = DESPAIR,
             color = Chocolate,
             shape = Chocolate  #<6>
            )
         ) +
  geom_point() +
  xlim(0, 1) +
  ylim(0, 1) +
  theme_bw() +
  labs(x = "Fraction of respondents who reported JOY",
       y = "Fraction of respondents who reported DESPAIR") +
  theme(text = element_text(size = 14))
```

6. Specify that the shape of the points should also display the chocolate/not-chocolate categorical variable.

##### Python code

```{python}
#| label: iter4Python
#| warning: false
#| message: false
#| code-fold: show
#| code-summary: "Python code for using color and shape together"

shapes = np.where(colors == 1, "^", "o")
just_candy_props['shapes'] = pd.Series(shapes, index = just_candy_props.index)

mask_choc = np.where(shapes == "^", True, False)
mask_nc = ~mask_choc

vmin = colors.min()
vmax = colors.max()

mpl.rcParams["font.size"] = 14

plt.style.use('seaborn-v0_8-deep')
fig, ax = plt.subplots()
scatter_choc = ax.scatter(just_candy_props['likeness'][mask_choc],just_candy_props['dislikeness'][mask_choc], c = just_candy_props['colors'][mask_choc], marker = "^", vmin = vmin, vmax = vmax, label = "Yes")
scatter_nc = ax.scatter(just_candy_props['likeness'][mask_nc],just_candy_props['dislikeness'][mask_nc], c = just_candy_props['colors'][mask_nc], marker = "o", vmin = vmin, vmax = vmax, label = "No")
ax.set_xlim(0,1)
ax.set_ylim(0,1)
ax.minorticks_on()
ax.grid(which='major', linestyle='-', linewidth='0.5', color='grey', alpha=0.7)
ax.grid(which='minor', linestyle=':', linewidth='0.3', color='grey', alpha=0.5)
ax.set_xlabel('Fraction of respondents who reported JOY')
ax.set_ylabel('Fraction of respondents who reported DESPAIR')


ax.legend(title="Chocolate")
plt.show()
plt.close()

```

:::
